---
title: "Runtime"
output: rmarkdown::html_vignette
bibliography: '`r system.file("REFERENCES.bib", package="SDModels")`'
vignette: >
  %\VignetteIndexEntry{Runtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

For this package, we have written methods to estimate regressions trees and random forests to minimize the spectral objective:

$$\hat{f} = \text{argmin}_{f' \in \mathcal{F}} \frac{||Q(\mathbf{Y} - f'(\mathbf{X}))||_2^2}{n}$$
The package is currently fully written in @RCoreTeam2024R:Computing for now and it gets quite slow for larger sample sizes. There might be a faster cpp version in the future, but for now, there are a few ways to increase the computations if you apply the methods to larger data sets.

## Computations

Some speedup can be achieved by taking advantage of modern hardware.

### Multicore

When estimating an SDForest, the most obvious way to increase the computations is to fit the individual trees on different cores in parallel. Parallel computing is supported for both Unix and Windows. Depending on how your system is set up, some linear algebra libraries might already run in parallel. In this case, the speed improvement from choosing more than one core to run on might not be that large. Be aware of potential RAM-overflows.

```{r core, eval=FALSE}
# fits the individual SDTrees in parallel on 22 cores
fit <- SDForest(x = X, y = Y, mc.cores = 22)

# performs cross validation in parallel
model <- SDAM(X, Y, cv_k = 5, mc.cores = 5)
```


## Approximations

In a few places, approximations perform almost as well as if we run the whole procedure. Reasonable split points to divide the space of $\mathbb{R}^p$ are, in principle, all values between the observed ones. In practice and with many observations, the number of potential splits grows too large. We, therefore, evaluate maximal `max_candidates` splits of the potential ones and choose them according to the quantiles of the potential ones.

```{r candidates, eval=FALSE}
# approximation of candidate splits
fit <- SDForest(x = X, y = Y, max_candidates = 100)
tree <- SDTree(x = X, y = Y, max_candidates = 50)
```

If we have many observations, we can reduce computing time by only sampling `max_size` observations from the data instead of $n$. This can dramatically reduce computing time compared to a full bootstrap sample but could also decrease performance.

```{r subsample, eval=FALSE}
# draws maximal 500 samples from the data for each tree
fit <- SDForest(x = X, y = Y, max_size = 500)
```